---
title: "Vectorising over lists with `map()` and friends"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(lobstr)
library(purrr)
library(palmerpenguins)
library(stringr)

penguins <- map(as.list(penguins), as.list)
```

## Intro

For this tutorial I've broken `penguins` again, this time turning it into a list of lists: 

```{r penguins_look}
tree(penguins, max_depth = 1)
```

In this tutorial we're going to be using that list-of-lists to get into the details of how functions like `map`, `map2`, `walk` and `pmap` work.

## `map()`

You've seen map before; it takes a vector and applies a function to each element, returning a list. `map()`is almost exactly equivalent to `lapply()` - the only real difference is that it will give more informative error messages.

You know the drill - we pass a vector and a function to `map()` and it does the rest.

```{r map_example}
increase_food_intake <- 
  function(weight) {
    random_addition <- rnorm(1, mean = 40, sd = 10)
    weight + random_addition
  }

## map(x, .f())
map(penguins$body_mass_g, increase_food_intake) |> 
  head(3)
```

### Side-bar: Using `...` with `map()`

As well as `x` and `.f`, `map()` will take `...`. Anything specified here is passed to `.f()` for each element. E.g.:

```{r dots_example}
add <- function(a, b) a + b

map(1:3, add, b = 3)
```

This can be a really useful syntax for writing code quickly and consicely. The people who make `map()` say:

> We now generally recommend against using ... to pass additional (constant) arguments to .f. Instead use a shorthand anonymous function. This makes it easier to understand which arguments belong to which function and will tend to yield better error messages.

That's been their recommendation ever since shorthand anonymous (i.e. `\(x){ }`) function syntax was added to base R, but I still sometimes use `...`. Use your judgement!

Have a go below at expressing my code above using equivalent shorthand anonymous function syntax. This will involve creating an anonymous function that wraps-around my `add()` function in the call to `map()`.

```{r express_shorthand, exercise = TRUE}
add <- function(a, b) a + b

map(1:3, )
```

## `map()` suffix variants

`map()` has four basic suffix variants: `map_lgl()`, `map_int()`, `map_dbl()` and `map_chr()`. These strictly return logical, integer (numeric), double (numeric) and character vectors respectively. Because they return vectors, they also impose additional constraints on `.f`, which must only return length-1 vector of the type corresponding to the suffix.

```{r map_suffix_quiz, echo = FALSE}
question(
  "Which of these calls won't crash?
* A: `map_lgl(penguins$species, str_to_lower)`,
* B: `map_dbl(penguins$bill_length_mm, \\(z) z / 2)`
* C: `map_int(penguins$body_mass_g, \\(x) x + 0.732)`
* D: `map_int(penguins$bill_depth_mm, \\(x) c(x, x + 2, x - 2)`",
answer("A"), answer("B", correct = TRUE), answer("C"), answer("D"),

correct = "Brilliant. A and B don't return the right vector type, and D does not return a length-1 vector. Note that `map_int()` is much stricter with integers than base R, which just turns them into doubles as soon as they stop being whole numbers.",
incorrect = "Remember that vector types must be consistent and that .f must return a vector with length 1",
allow_retry = TRUE
)
```

### Do it yourself

Groovy, now you can have a go at mapping with suffixes. Use a map-family function to map across `penguins$sex`, returning `TRUE` when a penguin is female and false when a penguin is male. Don't worry about `NA` values. Then pipe the output into `head()` so that we don't have to look at all 600 results.

```{r genderifier, exercise = TRUE}
map____() |> 
  head()
```

```{r, genderifier-solution, echo = FALSE}
## Lots of ways to do this, here is one:
is_female <- \(x) x == "female"

map_lgl(penguins$sex, is_female) |> 
  head()
```

## `map2()`

`map2()` is `map()` but with two lists in parallel. `map()`'s syntax is `map(.x, \(x) ...)` whereas `map2()`'s is `map2(.x, .y, \(x, y) ...)`. For example:

```{r map2_example}
## A function that takes two arguments:
estimate_bill_volume <- function(bill_depth, bill_length){
  assumed_bill_width <- 15
  bill_depth * bill_length * assumed_bill_width
}

## Mapped over two lists of equal length
map2(penguins$bill_depth_mm, penguins$bill_length_mm, estimate_bill_volume) |> 
  head()
```

As noted above, `.x` and `.y` must have equal lengths. `map2()` also has suffixed variants analogous to those of `map()`.

Use `map_dbl()` to calculate the bill_length/body_mass ratio of every penguin:

```{r map2_exercise, exercise = TRUE}
map2_dbl() |> 
  head()
```

```{r map2_exercise-solution, echo = FALSE}
map2_dbl(penguins$bill_length_mm, penguins$body_mass_g, \(x,y) x/y) |> 
  head()
```

## `pmap()`

`pmap()` is the final boss of mapping, and it allows you to map over an arbitrary number of vectors in parallel. Here's an exapmle that's mostly lifted from the function's documentation:

```{r pmap_example}
x <- list(1, 1, 1, 1)
y <- list(10, 20, 30, 40)
z <- c(100, 200, 300, 400)

pmap(list(x, y, z), sum)
```

Note that the vectors passed to `.f` are wrapped in a list. When `.f` has named arguments, the names of that list are used to specify which vector goes where:

```{r another_pmap_example, collapse = TRUE}
minus <- function(from, subtract){
  from - subtract
}

pmap_int(list(from = 1:5, subtract = 1), minus)
pmap_int(list(from = 1, subtract = 1:5), minus)
```

You might have noticed that I used some length-one vectors in that example - I haven't mentioned it yet but `map2()` and `pmap()`support recycling length-1 vectors.

### Exercise

This is slightly contrived, but hopefully helpful: Imagine that `vroom_write()` and `write_csv()` don't exist. In fact, nobody has thought to create a package or function that enables the writing of comma-separated values. We want to save `penguins` (currently a list-of-lists) as a comma-separated table, so we'll need to reformat it ourselves.

Use `pmap_chr()` and `str_c()` to convert our list-of-lists into a single character vector, where each element has the format "[species],[island],[bill_length_mm]...". For example, the first entry should look like this:
`"Adelie,Torgersen,39.1,18.7,181,3750,male,2007"`. 

Remember that `str_c()` takes `...`, so every element of `penguins` will be passed to `...` (since no elements share names with any of `str_c()`'s named arguments, `sep` and `collapse`)

```{r pmap_challenge, exercise = TRUE}
csv_vector <- 
  pmap_chr(
    #...
  )

head(csv_vector)
```


```{r pmap_challenge-solution}
csv_vector <- 
  pmap_chr(
    penguins,
    str_c,
    sep = ","
  )

head(csv_vector)
```

## Excellent bonus facts

As is often the case, I've learned some really nifty stuff whilst putting this tutorial together. Well, only really one nifty thing, which is this: If you supply a character, integer of list to `.f` in `map()`, those are treated as arguments to `pluck()` by default. So these calls are the same:

```{r cool_trick, collapse=TRUE}
map(penguins, 10)
map(penguins, \(x) pluck(x, 10))
```

This enables concise mapping for complicated lists:

```{r complicated_lists}
complicated_list <- 
  list(
    jimmy = list(height = 10, fruit = "apple"),
    susan = list(height = 8,  fruit = "banana"),
    lucy  = list(height = 14)
  )

# Are these 
map(complicated_list, "height") |> 
  map_lgl(\(x) x >= 10)
```
