---
title: "Indexing and assigning-into lists with purrr::pluck()"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=TRUE}
library(learnr)
library(purrr)
library(lobstr)
library(randomNames)
library(janitor)
library(stringr)

penguins <- readRDS("penguins-listified.RDS")
```

## Lists: a recap

### Properties of lists

#### Lists are a vector

You already know a lot about lists. They are technically a vector:

```{r list_is_vector, collapse = TRUE}
vector("list", length = 3L)
```


But they're not an _atomic_ vector; they don't behave like other vectors. For a start, lists can contain other vectors - even other lists:

```{r lists_contain_things, collapse=TRUE}
cool_list <- 
  list(
  four_twos = c(2, 2, 2, 2),
  
  some_numbers =list(c(11, 12, 13),
                     list(100, c(1000, 2000)))
)

lobstr::tree(cool_list)
```

They can also contain multiple types of data - something atomic vectors can never do:

```{r lists_contain_multiple_types, collapse = TRUE}
list(
  c("a", "b", "c"),
  TRUE,
  1:3,
  factor(c("dogs", "cats")),
  data.frame(a = 1:2, b = c("y", "z"))
  )
```

In short, lists are much less constrained than other data structures. This can make them very useful, but also very confusing to work with.

### Bonus fact!

**A data frame is a list of vectors.**

I'll keep saying this until you start to dream about it. From now when you see me don't say "Good morning dave" just say "A data frame is a list of vectors".

This also holds for tibbles, since those are an extension of dataframes. Don't believe me? I'll prove it:

We can make a list:

```{r df_is_list_of_vect}
listo <- list(
  fruit = c("mango", "apple", "clementine"),
  tier = ordered(c("S", "B", "D"), levels = c("S", "B", "D")) # (<< an ordered factor)
)
```

Give it a class and some row names...

```{r row_names}
class(listo) <- "data.frame"
row.names(listo) <- 1:3

```

And presto!

```{r presto}
listo
```

## Tiny additional recap: subsetting lists

You can subset lists in three ways:

* With a vector (of names or numerical indeces)
* With `[[ ]]` to unlist one element
* With `$` to unlist one element (watch out for partial matching!)

```{r subsetting_lists, collapse=TRUE}
foo <- list(first = TRUE, second = FALSE, third = TRUE, fourth = FALSE)

foo[2:3]
foo[[4]]
foo[["first"]]
foo$fou # note partial matching - more trouble than it's worth!
```

## Exploring lists interactively

Because lists can be infinitely nested into one another, you'll often want to get a high-level view of the overall structure of a list. Lets practice doing that with a relatively horrible list I've made by mutilating the `palmerpenguins` dataset. (On the upside, I've given all the penguins names 😊)

### `lobstr::tree()` is really good

My go-to function for list-investigation is `lobstr::tree()`. Have a go at it below, and tweak the `max_depth` and `max_length` arguments to see how the output varies.

```{r, tree_intro, exercise = TRUE}
tree(penguins,
     max_depth = 1,
     max_length = 5)
```

### `lengths()` is also great

`lengths()` is a base function that's equivalent to `map_dbl(x, length)`. Try getting the `lengths()` of `penguins` below:

```{r penguin_lengths, exercise = TRUE}

```

```{r penguin_lengths-solution}
lengths(penguins)
```

### `purrr::pluck_depth()` is also cool

`pluck_depth()` tells you how many levels you can index into an object:

```{r, pluck_depth_example, collapse=TRUE}
pluck_depth(1:3)
pluck_depth(list(1:3))
pluck_depth(list(list(1:3)))
```
Use `pluck_depth()` below to determine how many levels we could index into `penguins`:

```{r pluck_depth_trivial, exercise = TRUE}

```

```{r pluck_depth_trivial-solution}
pluck_depth(penguins)
```

## `pluck()`ing

`pluck()` is a `purrr` function that makes it very convenient to get deeply-nested list elements by name or index. For example, here is how we might find the sex of the Adelie penguin named Alifonso Chavez:

```{r}
pluck(penguins, "Adelie", "Alifonso_Chavez", "sex")
```

Now you try! Find me the bill depth of the Gentoo penguin named Rosa Williams:

```{r pluck_basics, exercise = TRUE}
pluck(penguins, )
```

```{r pluck_basics-solution}
pluck(penguins, "Gentoo", "Rosa_Williams", "dimensions", "bill_depth_mm")
```

### Pluck assignment

You can also use pluck to assign values to a list with `pluck()<-`:

```{r pluck_assign_example}
my_list <- list(
  cat_names = c("Cid", "Fluffy"),
  dog_names = c("Rex", "Fido")
)

pluck(my_list, "cat_names", 2) <- "mittens"
pluck(my_list, "cat_names")
```

One of the penguins, an Adelie named Merry Mcgill, has put on a little weight. Can you use `pluck()` to increase her `body_mass_g` by 200g? Check the hints if you're not sure.

```{r weight_increase, exercise = TRUE}

```

```{r weight_increase-hint-1}
pluck(penguins, ) <- 
```

```{r weight_increase-hint-2}
pluck(penguins, "Adelie", ) <- pluck(penguins, "Adelie", )
```

```{r weight_increase-solution}
pluck(penguins, "Adelie", "Merry_Mcgill", "dimensions", "body_mass_g") <-
  pluck(penguins, "Adelie", "Merry_Mcgill", "dimensions", "body_mass_g") + 200
```

#### Assigning `zap()`

There's a weird, inconsistent quirk in the way that R handles lists, and `purrr` implements a function called `zap()` to deal with it.

One of the differences between lists and atomic vectors is that lists can contain `NULL` elements:

```{r null_list, collapse = TRUE}
x <- list(a = 1, b = NULL)

x
```

So far so good. But the weird part is this: Assigning `NULL` to a list element makes that element dissapear completely!

```{r assign_null_to_list, collapse = TRUE}
x$a <- NULL
x

# x$a has dissapeared!
```

There is a way around this in base R: to assign `NULL` to `x$a` we would have to do `x["a"] <- list(NULL)`. But this is counterintuitive and inconsistent. Furthermore, it only works with `[` and not `$`. Not great.

By default, `pluck()` removes this odd behaviour and behaves as you would expect:

```{r pluck_assing_null, collapse = TRUE}
animal_cuteness <- list(cats = 10, bees = 9, wasps = 3)

pluck(animal_cuteness, "wasps") <- NULL
animal_cuteness

```

That's great! But what if we want to completely remove one element of a list? This is where `zap()` comes in:

```{r pluck_zap, collapse = TRUE}
pluck(animal_cuteness, "wasps") <- zap()
animal_cuteness
```

Honestly quite a fun function to use. Have a go in the exercise below:

One of the penguins, a Chinstrap by the name of Miranda Lanoue, has been found guilty of warcrimes. `zap()` her:

```{r warcrimes, exercise = TRUE}
pluck() <-
```

```{r warcrimes-solution}
pluck(penguins, "Chinstrap", "Miranda_Lanoue") <- zap()
```

## Plucking quiz

```{r quiz-1, echo = FALSE}
quiz(caption = "",
     question(
       "Two these expressions are equivalent. Which are they?
* A: `pluck(foo, \"index\", NULL)`
* B: `pluck(foo, \"index\") <- NULL`
* C: `pluck(foo, \"index\") <- zap()`
* D: `foo[[\"index\"]] <- NULL`",
       answer("A"), answer("B"), answer("C", correct = TRUE), answer("D", correct = TRUE)
     ),
     question(
       "What is the `pluck_depth()` of `list()`",
       answer("0"), answer("1", correct = TRUE), answer("2")
     ),
     question(
       "What is the `pluck_depth()` of `list(1:10)`",
       answer("0"), answer("1"), answer("2", correct = TRUE)
     ),
     question(
       "What is the `pluck_depth()` of `list(\"hello\")`",
       answer("0"), answer("1"), answer("2", correct = TRUE)
     ))
```

